Draft Article

### 2019-12-24 article.tex
Remove extra comments aerial ending

% Farquhar cites this as a landmark moment in the history of aerial reconnaissance. \enquote{No longer would ferret operations be conducted ad hoc by the military services;} he writes, \enquote{from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S. foreign policy.}\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the \enquote{serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.}\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: \enquote{Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.}\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.

@@

Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident. He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany. And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with... a diplomatic protest. The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}

% The U-2 did not disappear after the Powers flight. A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela. Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded.\footcite[p.~195]{pedlow_central_1992} As it turned out, it never became necessary to resume Soviet Bloc overflights because satellite photo-reconnaissance soon replaced them; U-2 overflights were rendered mostly obsolete. There is, however, one use for which satellites will likely never displace aerial photo-reconnaissance---a situation like the Cuban Missile Crisis, where events are so time-sensitive that intelligence is needed in a matter of hours.

% On October 15, 1962, U-2 photography revealed that the Soviet Union was assembling missiles in Cuba that were capable of carrying nuclear warheads. As the crisis unfolded, the United States continued to send U-2 planes over Cuba in order to maintain a close watch on the development of the missile sites. Kennedy's team decided that if a U-2 were to be shot down, the most likely SAM site would be immediately fired upon.\footcite[p.~217]{caro_passage_2013} Then, on October 27, it happened---a U-2 was shot down flying over Cuba and the pilot was killed. Kennedy flinched; he had just sent Khrushchev an offer that might lead to peace, and he didn't want to jeopardize it. Against the advice of administration hard-liners, including Vice President Johnson, Kennedy refused to authorize the strike against the Cuban SAM site.\footcite[p.~220]{caro_passage_2013} The next day, Khrushchev agreed to Kennedy's terms, and the Cuban Missile Crisis ended with only a single casualty---the U-2 pilot, Major Rudolph Anderson.

% As his title indicates, Rudolph Anderson was an Air Force Officer, not a member of the CIA. During the crisis, the President transferred control of Cuban U-2 reconnaissance to the Air Force, and authorized it to fly as many missions as necessary over the island.\footcite[p.~208-209]{pedlow_central_1992} Khrushchev did not say much about the Cuban U-2 in his writing, but he was very clear on one point: he did not order the SAM strike, and was worried that Kennedy \enquote{might not be able to absorb this blow.}\footcite[p.~338]{khrushchev_memoirs_2007} Shooting down an American aircraft risked escalation, and that was not something Khrushchev wanted. This was the second time in history that Soviet forces had shot down a U-2 plane---but even though the spy plane was flying over Cuban airspace and piloted by an US Air Force Major, it was the Soviets who were worried that shooting it down had been an unnecessarily aggressive act.

% Even when a U-2 operation went poorly, it still played by the rules of the game. Powers opted against using his suicide capsule, but the macabre truth about the U-2 is that pilots weren't supposed to survive a crash at all. The United States originally publicized the weather craft story because Eisenhower had been assured that Gary Powers could not possibly be alive to tell the Russians the truth.\footcite[p.~35]{lindgren_trust_2000} When Powers confessed to his Soviet captors, he told them, correctly, that he was a civilian pilot with the CIA. His cover had been blown, but underneath it he was still a spy, and in pretty much all respects he and the U-2 were treated accordingly.

>>>>

@@

Until the day he died, Khrushchev greatly exaggerated the lasting effect of the U-2 Incident. He wrote in his memoir that \enquote{the commander of American forces in West Germany gave the order not to fly any closer than 50 kilometers from the border between East and West Germany. And no more incidents of that kind occurred.}\footcite[p.~256]{khrushchev_memoirs_2007} His son Sergei, who annotated the memoir, admitted that this statement was false: \enquote{In practice such incidents occurred again from 1961 onward, but instead of the U-2 the Americans now used various types of Phantom or SR-71.}\footcite[p.~258]{khrushchev_memoirs_2007} The United States continued to use the U-2 extensively in other regions as well.\footnote{A recently declassified CIA report details the many locations in which U-2 photography was employed to gather intelligence after 1960, including China, India, Indonesia, Thailand, Tibet, Laos, North Vietnam, and Venezuela. Lyndon Johnson even clarified that his predecessor's order to end U-2 flights over the Soviet Bloc was not indefinitely binding---it was valid only until countermanded. For more, see: \cite[p.~195]{pedlow_central_1992}}. On at least two occasions after the U-2 Incident, an American overflight crossed over into Soviet territory and the Soviets responded with... a diplomatic protest. The US quickly apologized, and that was that.\footcite{orlov_u-2_2007}
